## Grouped Barplot in R

In this post, we are going to see how we can program a grouped barplot in R. We will be using R to explore data related to bike share systems for three major cities in the United States—Chicago, New York City, and Washington.


We will use data provided by Motivate, a bike-share system provider for many major cities in the United States, to answer a question about the data.


We will explore the question, "What are the counts for each user type?".

## The Dataset

Let's import our datasets and see what we are working with.


```{r}
ny = read.csv("new_york_city.csv")
wash = read.csv("washington.csv")
chi = read.csv("chicago.csv")
```

We now use the head function to see how each data frame looks like.

```{r}
head(ny)
```

```{r}
head(wash)
```

```{r}
head(chi)
```

## Importing The Needed Libraries

To accomplish our aim, we need just two libraries. ggplot2 and dplyr. If you don't have these, install them as follows.

```{r}
library(ggplot2)
library(dplyr)
```

If you don't have these libraries, use this to install
```{r}
if (!require('ggplot2')) {
  install.packages("ggplot2")
}
if (!require('dplyr')) {
  install.packages("dplyr")
}
```
 
## Getting the counts of different user types for each city

The dplyr library provides the count() function for just that. We use the %>% operator to pipe the data frame into the count function as follows.

```{r}
# We want to count frequency of each User.Type
(ny_user_types <- ny %>% count(User.Type))
(chi_user_types <- chi %>% count(User.Type))
(wash_user_types <- wash %>% count(User.Type))
```

We wrapped each expression with "()" to run it immediately.
Notice something about the output? We get an empty User.Type with count of 1. Let's filter that out by slicing as follows.

```{r}
(ny_types <- ny_user_types[2:nrow(ny_user_types), ])
(chi_types <- chi_user_types[2:nrow(chi_user_types), ])
(wash_types <- wash_user_types[2:nrow(wash_user_types), ])
```
R indexing starts from 1 so for the number of rows we want, we slice from the second row, that is, 2 to the total number of rows in the data frame with the nrow() function. We leave the number of columns empty since we want all of it.

## Combining our User.Type counts.

For us to be able to compute the grouped barplot, we need to map the User types and their counts to the cities the belong in a data frame.
The functions we will be using are the
The rep() function: We use this to repeat a vector of cities.
```{r}
(cities <- rep(c("Washington" , "New York" , "Chicago") , each=2)) 
# We want to repeat each city twice since there are two user types: Subscriber and Customer
```

The rbind() function: We use this to vertically combine the User.Type data frames above into one.
```{r}
(data <- rbind(wash_types, ny_types, chi_types))
```

Finally, we map the User.Type counts to their respective cities.

```{r}
(data.m <- data.frame(data, cities))
```
Now we have our data in a format we can work with to produce our barplot.

## Plotting our data frame

For our plot, we want to plot the counts on the y-axis and the user types on the x-axis.

```{r}
ggplot(data.m, aes(fill=cities, y=n, x=User.Type)) + geom_col(position="dodge") + labs(title="User Type Counts for Cities")
```

From the above visualizations:

 1. For New York, there are more subscribers than customers.
 2. For Chicago, there are more subscribers than customers.
 3. For Washington, there are more subscribers than customers.

 Hence, more Subscribers use the service than customers

It's as simple as that.

## A Little sidenote

When I was practicing how to program grouped bar plots, I didn't have access to the dplyr library in the environment I was working on, so I wrote this little function to compute the counts for each user type.

```{r}
user_counts.function <- function(city_data) {
    aggregated_data <- aggregate(city_data$User.Type, by=list(city_data$User.Type), FUN=length)
    names(aggregated_data)[1] = "User.Type"
    names(aggregated_data)[2] = "Count"
    row.names(aggregated_data) <- NULL
    return (aggregated_data[2:nrow(aggregated_data), ])
}
```
We use the aggregate function and apply length function to aggregate by the User.Type column of the city_data parameter.
Note: Use this only if for any reason you do not have access to the dplyr library.

Hope you enjoyed this post. Leave a comment below!